# Analyzing and visualizing data {#datanalysis}

----

```{r libraries_elegans, include=FALSE}
library(readxl)
library(tidyverse)
library(toolboxr)
library(patchwork)
library(cowplot)
library(here)
```

The data that is to be visualized comes from an experiment involving C. elegans nematodes and was supplied by J. Louter of the INT/ILC. In this experiment, the number of offspring of adult C. elegans was counted after exposing them to different substances in multiple concentrations.

## Preparing the data

### Inspecting the data in R

```{r import data}
celegans_data <- read_excel(here("data_raw/CE.LIQ.FLOW.062_Tidydata.xlsx"))
```

Before the data can be visualized, it has to be cleaned up. The dataframe is already in tidy format, but the data types of the columns need to be changed. Expected data types would be double/integer for RawData (number of offspring), character for compName (name of the compound) and double for compConcentration (concentration of the compound). The actual data types of the columns are double for RawData, character for compName and character for compConcentration. The data type for compConcentration has not been correctly assigned which will cause the following issue if it's not changed:

```{r alphabeticalplot, warning=FALSE, fig.cap="Number of C. elegans offspring under a number of circumstances where the x-axis is ordered alphabetically"}
#Create a scatter plot
plot_chr <- ggplot(data = celegans_data, aes(x = compConcentration, y = RawData))+
  geom_point(aes(color = compName,
                 shape = expType),
             size = 1)+
  labs(x = "Concentration",
       y = "Number of offspring",
       title = "Number of C. elegans offspring under\na number of circumstances, alphabetical",
       shape = "Type",
       color = "Compound")+
  scale_colour_manual(values = c("red", "darkgoldenrod1", "green", "royalblue3", "violet"))+
  theme_minimal()+
  theme(legend.key.size = unit(0.75,"line"),
        legend.text = element_text(size = 8),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plot_chr
```

The x-axis labels are ordered alphabetically because the data type of the compConcentration is character instead of double. R probably read it this way because Excel adds an E to showcase exponential values.

### Fixing the character/double issue

Since not all compounds share the same compUnit (the unit in which their concentration is measured), 2 separate plots need to be made and then combined: one for the compounds measured in nM and one for the compounds measured in percentage. After changing the data type from character to double and adding jitter to the plot to make it more readable it looks as follows:

```{r scatterplotcelegans, warning=FALSE, fig.cap="Total number of C. elegans offspring under a number of circumstances, with on the y-axis the number of offspring and on the x-axis the log10 concentration in nM."}
#Change data type from chr to dbl
celegans_data$compConcentration <- as.double(celegans_data$compConcentration)

#Create a scatter plot for the concentration in nM
celegans_data_nM <- celegans_data %>% filter(compUnit == "nM")
plot_nM <- ggplot(data = celegans_data_nM, aes(x = log10(compConcentration), y = RawData))+
  geom_jitter(aes(color = compName, #add jitter
                  shape = expType),
              width = 0.5, height = 0.2)+
  labs(x = "log10 concentration (nM)",
       y = "Number of offspring")+
  coord_cartesian(ylim = c(0, 120))+
  scale_shape_manual(values = 3)+
  scale_colour_manual(values = c("red", "darkgoldenrod1", "royalblue3"))+
  theme_minimal()+
  theme(legend.position = "none")

#Create a scatter plot for the concentration in pct
celegans_data_pct <- celegans_data %>% filter(compUnit == "pct")
plot_pct <- ggplot(data = celegans_data_pct, aes(x = expType, y = RawData))+
  geom_jitter(aes(color = compName, #add jitter
                  shape = expType),
              size = 1.3,
              width = 0.1, height = 0.075)+
  xlab("Type")+
  coord_cartesian(ylim = c(23.7, 121))+
  scale_colour_manual(values = c("green", "violet"))+
  theme_minimal()+
  theme(legend.position = "none",
        axis.text.x=element_text(vjust=0.5, hjust=0.5, size = 8.75),
        axis.text.y = element_blank(), 
        axis.ticks.y = element_blank(), 
        axis.title.y = element_blank())+ #remove any y-axis labeling
  rotate_axis_labels("x", 90)

#obtain legend
legend <- get_legend(plot_chr)

#combine figures and legend for readability
ggdraw(plot_grid(plot_grid(plot_nM, plot_pct),
                 plot_grid(NULL, legend, ncol = 3),
                 rel_widths = c(1, 0.35)))+
  plot_annotation("Number of C. elegans offspring under\na number of circumstances")
```

The positive control for this experiment is ethanol. The negative control for this experiment is S-medium.

## Analyzing and visualizing the data

### Plan for IC50/ED50 visualization

Since this chapter is focused on visualization, there won't be a statistical analysis of the dose-response interactions. If I were to perform a statistical analysis, it would be done using the {drc} package [@ritzDoseResponseAnalysisUsing2015], which has functions to calculate IC50s and ED50s, which then could be compared. Using {ggplot2} [@wickhamGgplot2ElegantGraphics2016], a scatter plot can be made with on the y-axis the number of offspring and on the x-axis the compound concentration. The calculated IC50/ED50 values could be added as lines over the plot, giving every compound (line) a different color. An ANOVA can also be conducted to look for differences between concentrations, after checking if the data comes from a normal distribution using the Shapiro-Wilk test.

### Normalizing the data
To make is easier to read the results visualized in the plot, the number of offspring was normalized as a fraction of the negative control, meaning any value under 1 shows a reduction in the number of offspring. To normalize the data, the following code was used:

```{r normalization}
#Obtain the mean of the RawData (negativeControl)
controlNeg <- celegans_data %>% filter(compName == "S-medium") %>% summarize(mean = mean(RawData, na.rm = TRUE))

#Use the mean to calculate fractions
mutated <- celegans_data %>% filter(RawData > 0) %>% 
  select(RawData, compName, compConcentration, expType) %>% na.omit() %>% 
  mutate(normalized = RawData/controlNeg$mean)

#Obtain the means of the other 2 control groups
controlPos <- mutated %>% filter(expType == "controlPositive") %>% summarize(mean = mean(normalized, na.rm = TRUE))
controlVeh <- mutated %>% filter(expType == "controlVehicleA") %>% summarize(mean = mean(normalized, na.rm = TRUE))
```

### Creating a legend

To give the final figure a proper legend, the legend needs to be generated in a separate figure since we don't want to include the test-compounds in it. The following code was used to generate a legend, making use of dummy data.

<details>

<summary>Generating a legend using dummy data</summary>

```{r dummylegend}
dummydf <- data.frame(a = rep(c(1, 2, 3), each=2), #create dummy dataframe for the dummy plot
                      b = c(1, 4, 6, 3, 4, 7),
                      Control = rep(c("controlNegative", "controlPositive", "controlVehicleA"), each=2))

dummyplot <- ggplot(dummydf, aes(x=a, y=b, group=Control))+ #create dummy plot to obtain its legend
  geom_line(aes(linetype=Control, color=Control))+
  scale_linetype_manual(values=c("solid", "solid", "longdash"))+
  scale_color_manual(values=c("violet", "green", "green"))

dummylegend <- get_legend(dummyplot) #obtain dummy legend
```

</details>

### Generating the normalized plot

With the data normalized and the legend obtained, the normalized plot can be generated.

```{r normalizedplot, fig.cap="Number of C. elegans offspring as a fraction of the negative control group, with on the y-axis the normalized number of offspring and on the x-axis the log10 concentration of the compounds in nM."}
#create normalized plot
plotfraction <- mutated %>% filter(compName == "2,6-diisopropylnaphthalene" | compName == "decane" | compName == "naphthalene") %>%
  ggplot(aes(x = log10(compConcentration), y = normalized))+
  geom_jitter(aes(color = compName), width = 0.5, height = 0.1)+
  scale_colour_manual(values = c("red", "darkgoldenrod1",  "royalblue3"))+
  coord_cartesian(ylim = c(0, 1.5))+
  labs(x = "log10 concentration (nM)",
       y = "Normalized number of offspring",
       title = "Number of C. elegans offspring as a fraction\nof the negative control group",
       color = "Compound")+
  geom_hline(yintercept = 1, color = "violet")+
  geom_hline(yintercept = controlPos$mean, color = "green")+
  geom_hline(yintercept = controlVeh$mean, color = "green", linetype = "longdash")+
  theme_minimal()+
  guides(color = "none")+
  facet_wrap(~ compName)

#combine figure and legend
ggdraw(plot_grid(plotfraction, dummylegend,
       rel_widths = c(5, 1)))
```

This way it's easier to read the graph. Everything below the pink line means less offspring than control C. elegans and everything above it means more. The figure (\@ref(fig:normalizedplot)) shows a decrease in the number of offspring for all 3 groups.