# Analyzing data from an experiment {#datanalysis}

```{r setup, eval=FALSE, include=FALSE}
install.packages("readxl")
install.packages("devtools")
library(devtools)
devtools::install_github("uashogeschoolutrecht/toolboxr")
install.packages("tidyverse")
install.packages("patchwork") 
install.packages("cowplot")
```

```{r libraries_elegans, include=FALSE}
library(readxl)
library(tidyverse)
library(toolboxr)
library(patchwork)
library(cowplot)
```

In this experiment, the number of offspring of adult C. elegans was counted after exposing the nematodes to different substances in multiple concentrations. The data was supplied by J. Louter of the INT/ILC.

## Open the file in R and inspect the data.

```{r A}
celegans_data <- read_excel("/Users/nineluijendijk/Downloads/CE.LIQ.FLOW.062_Tidydata.xlsx")
```

Expected data types would be double/integer for RawData (number of offspring), character for compName (name of the compound) and double for compConcentration (concentration of the compound). The actual data types of the columns are double for RawData, character for compName and character for compConcentration. The data type for compConcentration has not been correctly assigned which will cause the following issue if it's not changed:

```{r alphabetical_plot, warning=FALSE, fig.cap="Number of C. elegans offspring under a number of circumstances where the x-axis is ordered alphabetically"}
#Create a scatter plot
plot_chr <- ggplot(data = celegans_data, aes(x = compConcentration, y = RawData))+
  geom_point(aes(color = compName,
                 shape = expType),
             size = 1)+
  rotate_axis_labels("x", 90)+
  labs(x = "Concentration",
       y = "Number of offspring",
       title = "Number of C. elegans offspring under\na number of circumstances, alphabetical",
       shape = "Type",
       color = "Compound")+
  scale_colour_manual(values = c("red", "darkgoldenrod1", "green", "royalblue3", "violet"))+
  theme(legend.key.size = unit(0.75,"line"),
        legend.text = element_text(size = 8))
plot_chr
```

The x-axis labels are ordered alphabetically because the data type of the compConcentration is character instead of double. R probably read it this way because Excel adds an E to showcase exponential values.

## Fixing the character/double issue

After changing the data type from character to double and adding jitter to the plot it looks as follows:

```{r , warning=FALSE}
#Change data type from chr to dbl
celegans_data$compConcentration <- as.double(celegans_data$compConcentration)

#Create a scatter plot for the concentration in nM
celegans_data_nM <- celegans_data %>% filter(compUnit == "nM")
plot_nM <- ggplot(data = celegans_data_nM, aes(x = log10(compConcentration), y = RawData))+
  geom_jitter(aes(color = compName, #add jitter
                  shape = expType),
              width = 0.5, height = 0.2)+
  labs(x = "log10 concentration (nM)",
       y = "Number of offspring")+
  coord_cartesian(ylim = c(0, 120))+
  scale_shape_manual(values = 3)+
  scale_colour_manual(values = c("red", "darkgoldenrod1", "royalblue3"))+
  theme(legend.position = "none")

#Create a scatter plot for the concentration in pct
celegans_data_pct <- celegans_data %>% filter(compUnit == "pct")
plot_pct <- ggplot(data = celegans_data_pct, aes(x = expType, y = RawData))+
  geom_jitter(aes(color = compName, #add jitter
                  shape = expType),
              size = 1.3,
              width = 0.1, height = 0.075)+
  xlab("Type")+
  coord_cartesian(ylim = c(23.7, 121))+
  scale_colour_manual(values = c("green", "violet"))+
  theme(legend.position = "none",
        axis.text.x=element_text(vjust=0.5, hjust=0.5, size = 8.75),
        axis.text.y = element_blank(), 
        axis.ticks.y = element_blank(), 
        axis.title.y = element_blank())+ #remove any y-axis labeling
  rotate_axis_labels("x", 90)

#obtain legend
legend <- get_legend(plot_chr)

#combine figures and legend for readability
ggdraw(plot_grid(plot_grid(plot_nM, plot_pct),
                 plot_grid(NULL, legend, ncol = 3),
                 rel_widths = c(1, 0.35)))+
  plot_annotation("Number of C. elegans offspring under\na number of circumstances")

```

The positive control for this experiment is ethanol. The negative control for this experiment is S-medium.

## Think about how you would analyze this experiment to learn whether there is indeed an effect of different concentrations on offspring count and whether the different compounds have a different curve (IC50). Write down you analysis as a step-wise plan.

## Normalize the data for the controlNegative in such a way that the mean value for controlNegative is exactly equal to 1 and that all other values are expressed as a fraction thereof. Rerun your graphs with the normalized data.

```{r J}
#Obtain the mean of the RawData (negativeControl)
controlNeg <- celegans_data %>% filter(compName == "S-medium") %>% summarize(mean = mean(RawData, na.rm = TRUE))

#Use the mean to calculate fractions
mutated <- celegans_data %>% filter(RawData > 0) %>% 
  select(RawData, compName, compConcentration, expType) %>% na.omit() %>% 
  mutate(normalized = RawData/controlNeg$mean)

#Obtain the means of the other 2 control groups
controlPos <- mutated %>% filter(expType == "controlPositive") %>% summarize(mean = mean(normalized, na.rm = TRUE))
controlVeh <- mutated %>% filter(expType == "controlVehicleA") %>% summarize(mean = mean(normalized, na.rm = TRUE))

mutated %>% filter(compName == "2,6-diisopropylnaphthalene" | compName == "decane" | compName == "naphthalene") %>%
  ggplot(aes(x = log10(compConcentration), y = normalized))+
  geom_jitter(aes(color = compName), width = 0.5, height = 0.1)+
  scale_colour_manual(values = c("red", "darkgoldenrod1",  "royalblue3"))+
  coord_cartesian(ylim = c(0, 1.5))+
  labs(x = "log10 concentration (nM)",
       y = "Normalized number of offspring",
       title = "Number of C. elegans offspring as a fraction\nof the negative control group",
       color = "Compound")+
  geom_hline(yintercept = 1, color = "violet")+
  geom_hline(yintercept = controlPos$mean, color = "green")+
  geom_hline(yintercept = controlVeh$mean, color = "green", linetype = "longdash")+
  guides(color = "none")+
  annotate("text", x = -3.8, y = 0.55, label = "controlPositive", size = 2)+
  annotate("text", x = -3.8, y = 0.95, label = "controlNegative", size = 2)+
  annotate("text", x = -3.8, y = 1.05, label = "controlVehicleA", size = 2)+
  facet_wrap(~ compName)
```

This way it's easier to read the graph. Everything below the pink line means less offspring than control C. elegans and everything above it means more.